---
title: "Pseudomonas Epiphytic Growth and Virulence Analysis"
date: "2024 Summer"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
  html_document:
    css: css/tufte.css
    toc: true
    toc_float: true
  word_document:
    toc: true
    reference_docx: templates/wordstyletemplate.docx
    keep_md: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy.opts = list, tidy = TRUE, fig.align = "center", # Always relative to the document directory
                      fig.path = "../figures/", # Send any figures to this folder
                      dev = "png",  # Export figures as PNG
                      dpi = 350)  # Set the resolution to 300 dots per inch (dpi)
```

# Load Libraries

```{r load-packages}
pacman::p_load(ggplot2, readxl, RColorBrewer, paletteer, viridis, wesanderson, ggbeeswarm, dplyr, tidyverse, devtools, emmeans, cowplot, knitr, survival, here, tibble, survminer, lubridate, formatR, gridExtra, ggsurvfit, gtsummary, tidycmprsk, install = FALSE)
```

# Load Data

```{r load-data}
# Define the desired order of the treatments
strains_order <- c("194", "200", "204", "205", "215", "216", "220", "221", "227", "228", "B728a", "Cit7", "pisi", "Control")

strains_df <- read_excel("data/strains_R.xlsx")

epi_growth_data <- read_excel("data/pseud_epi_growth_2024summer_R.xlsx")

# Convert the strain variable to a factor with the specified order
epi_growth_data$treatment <- factor(epi_growth_data$treatment, levels = strains_order)

# Join the strains dataframe to epi growth
epi_growth_data <- left_join(epi_growth_data, strains_df, by = "treatment")

# Load virulence data
aphid_virulence_data <- read_csv("data/virulence_new_all.csv")

# Convert the treatment variable to a factor with the specified order
aphid_virulence_data$treatment <- factor(aphid_virulence_data$treatment, levels = strains_order)

# Join the strains dataframe to virulence
aphid_virulence_data <- left_join(aphid_virulence_data, strains_df, by = "treatment")
```

# Load colors

```{r colors}
# Extract the default ggplot colors
default_colors <- scales::hue_pal()(length(strains_order) - 1)

# Create a named vector for strain colors, ensuring "Control" is black
strain_colors <- setNames(c(default_colors, "black"), strains_order)

# Load species colors
species_colors <- c(
  "P. fluorescens" = "dodgerblue2",
  "P. syringae" = "springgreen3",
  "P. putida" = "sienna4",
  "Control" = "black")

species_colors_manual <- c("194" = "sienna4", 
                           "200" = "dodgerblue2", 
                           "204" = "dodgerblue2", 
                           "205" = "dodgerblue2", 
                           "215" = "springgreen3", 
                           "216" = "dodgerblue2", 
                           "220" = "sienna4", 
                           "221" = "dodgerblue2", 
                           "227" = "dodgerblue2", 
                           "228" = "dodgerblue2", 
                           "B728a" = "springgreen3", 
                           "Cit7" = "springgreen3", 
                           "pisi" = "springgreen3", 
                           "Control" = "black")
```

# Goals

-   Create Kaplan-Meier curve for Pseud. virulence data
-   Use stats (Wilcox?) to determine statistical significance of each strain
-   Compare virulence data with epiphitic growth ability

# Virulence Analysis

Note: *In order to help me with this analysis, I am using the following sites - [Survival Analysis in R](https://www.emilyzabor.com/tutorials/survival_analysis_in_r_tutorial.html) and [Hazard Ratio: Interpretation & Definition](https://statisticsbyjim.com/probability/hazard-ratio/).*

## Caclulcate survival probabilities for each strain and create dataframe

### Survival model

```{r survival-model}
# Fit the survival model
km_fit <- survfit(Surv(time, censored) ~ treatment, data = aphid_virulence_data)
```

### Extract survival probability
```{r test-survival}
# Fit the survival model
km_fit <- survfit(Surv(time, censored) ~ treatment, data = aphid_virulence_data)

# Extract survival probabilities at specific time points
time_points <- c(72)
km_summary <- summary(km_fit, times = time_points)

# Calculate standard error (assuming fit includes the necessary information)
std_error <- summary(km_fit, times = time_points)$std.err

# Initialize empty lists to store the results
times_list <- list()
treatment_list <- list()
surv_prob_list <- list()
std_error_list <- list()

# Loop over each treatment group and extract survival probabilities at specified time points
for (i in 1:length(km_fit$strata)) {
  treatment_name <- names(km_fit$strata)[i]
  for (t in time_points) {
    idx <- which(km_summary$time == t & km_summary$strata == treatment_name)
    if (length(idx) > 0) {
      times_list <- c(times_list, t)
      treatment_list <- c(treatment_list, treatment_name)
      surv_prob_list <- c(surv_prob_list, km_summary$surv[idx])
      std_error_list <- c(std_error_list, km_summary$std.err[idx])
    } else {
      times_list <- c(times_list, t)
      treatment_list <- c(treatment_list, treatment_name)
      surv_prob_list <- c(surv_prob_list, NA)
      std_error_list <- c(std_error_list, NA)
    }
  }
}

# Create the data frame
surv_probs <- data.frame(
  time = unlist(times_list),
  treatment = unlist(treatment_list),
  surv_prob = unlist(surv_prob_list),
  std_error_data = unlist(std_error_list)
)

# Replace "treatment=" with an empty string
surv_probs$treatment <- gsub("treatment=", "", surv_probs$treatment)

# Convert the treatment variable to a factor with the specified order
surv_probs$treatment <- factor(surv_probs$treatment, levels = strains_order)

# Join the survival dataframe to strains
surv_probs <- left_join(surv_probs, strains_df, by = "treatment")

# Print the data frame
print(surv_probs)
```

### Post-hoc test for survival

```{r post-hoc}
# Perform pairwise log-rank tests between treatment groups and the control group
pairwise_tests_treatment <- pairwise_survdiff(Surv(time, censored) ~ treatment, data = aphid_virulence_data)

# Print the pairwise tests
print(pairwise_tests_treatment)

# Perform pairwise log-rank tests between species_complex groups and the control group
pairwise_tests_species <- pairwise_survdiff(Surv(time, censored) ~ species_complex, data = aphid_virulence_data)

# Print the pairwise tests
print(pairwise_tests_species)
```

### Plot Surival Dataframe at different times

```{r surival-barplot}
surv_plot <- ggplot(data = surv_probs, aes(x = treatment, y = (1-surv_prob), color = species_complex, fill = species_complex, alpha = 0.95)) +
  geom_col() +
  geom_errorbar(aes(ymin = (1-surv_prob) - std_error_data, ymax = (1-surv_prob) + std_error_data), 
                width = 0.3,  # Width of error bars
                position = position_dodge(width = 0.9)) +  # Dodge bars slightly
  labs(title = "72 hours",
       x = "-80 Strain #",
       y = "Aphid Mortality at 72 Hours (%)") +
  theme_minimal() +
  scale_fill_manual(values = species_colors) +
  scale_color_manual(values = species_colors) +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "bottom")

surv_plot
```

## Make Kaplan-Meier Plot

### Cohort Survival Curve

```{r Kaplan-Meier}
# Plot the Kaplan-Meier survival curves by treatment
surv_curv_treatment <- survfit2(Surv(time, censored) ~ treatment, data = aphid_virulence_data) %>%
  ggsurvfit(type = "survival") +
  labs(
    x = "Hours",
    y = "Overall survival probability"
  ) +
  theme_minimal() +
  scale_fill_manual(values = species_colors_manual) +
  scale_color_manual(values = species_colors_manual) +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1), 
        legend.position = "right")

# Plot the Kaplan-Meier survival curves by species complex
surv_curv_species <- survfit2(Surv(time, censored) ~ species_complex, data = aphid_virulence_data) %>%
  ggsurvfit(type = "survival") +
  labs(
    x = "Hours",
    y = "Overall survival probability"
  ) +
  theme_minimal() +
  scale_fill_manual(values = species_colors) +
  scale_color_manual(values = species_colors) +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1), 
        legend.position = "right")

#show plots
surv_curv_treatment
surv_curv_species
```

# Epiphytic Growth Analysis

## Calculate mean/variance epiphytic growth ability

```{r mean-epi-growth}
# Replace NA with a lower value or remove them for visualization
# Remove rows with NA in CFU_per_10_leafdiscs
epi_growth_clean <- epi_growth_data %>% 
  filter(!is.na(CFU_per_10_leafdiscs))

# Convert CFU_per_10_leafdiscs to numeric, handling scientific notation
epi_growth_clean$CFU_per_10_leafdiscs <- as.numeric(gsub("<", "", epi_growth_clean$CFU_per_10_leafdiscs))

# Subset data to remove unfinished strains and extract the letter part from the 'plant_rep' column
epi_growth_clean <- epi_growth_clean %>%
  mutate(plant_rep_letter = substr(plant_rep, 1, 1))

# Calculate the mean and SD CFU per strain for each plant rep letter and retain the specified columns
epi_growth_mean_per_rep <- epi_growth_clean %>%
  group_by(treatment, plant_rep_letter) %>%
  summarise(
    mean_CFU_per_10_leafdiscs = mean(CFU_per_10_leafdiscs, na.rm = TRUE),
    sd_CFU = sd(CFU_per_10_leafdiscs, na.rm = TRUE),
    sample = first(sample),
    person = first(person),
    species_complex = first(species_complex),
    block = first(block),
    treatment = first(treatment),
    plant_rep = first(plant_rep),
    .groups = 'drop'
  )

# Calculate the mean and SD CFU per strain for each plant rep letter and retain the specified columns
epi_growth_mean_per_strain <- epi_growth_clean %>%
  group_by(treatment) %>%
  summarise(
    mean_CFU_per_10_leafdiscs = mean(CFU_per_10_leafdiscs, na.rm = TRUE),
    sd_CFU = sd(CFU_per_10_leafdiscs, na.rm = TRUE),
    sample = first(sample),
    person = first(person),
    species_complex = first(species_complex),
    block = first(block),
    treatment = first(treatment),
    plant_rep = first(plant_rep),
    .groups = 'drop'
  )

epi_growth_mean_per_strain
```

## Plot Epiphytic Growth

```{r epi-growth}
# plot by treatment
epi_growth_treatment <- ggplot(data = epi_growth_mean_per_rep,
                               aes(x = treatment, 
                                   y = mean_CFU_per_10_leafdiscs,
                                   color = species_complex,
                                   fill = species_complex)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.3) +
  geom_beeswarm(stroke = 0.5, size = 0.8, alpha = 0.8) +
  theme_minimal() +
  scale_fill_manual(values = species_colors) +
  scale_color_manual(values = species_colors) +
  scale_y_log10() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")

# plot by species complex
epi_growth_species <- ggplot(data = epi_growth_mean_per_rep,
                               aes(x = species_complex, 
                                   y = mean_CFU_per_10_leafdiscs,
                                   color = species_complex,
                                   fill = species_complex)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.3) +
  geom_beeswarm(stroke = 0.5, size = 0.8, alpha = 0.8) +
  theme_minimal() +
  scale_fill_manual(values = species_colors) +
  scale_color_manual(values = species_colors) +
  scale_y_log10() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")


# show plots
epi_growth_treatment
epi_growth_species
```

# Epiphytic growth stats

```{r epi-growth-stats}
# stats per treatment
epi_growth_mod_treatment = lm(mean_CFU_per_10_leafdiscs ~ treatment, data = epi_growth_mean_per_rep)

emmeans(epi_growth_mod_treatment, pairwise ~ treatment)

hist(resid(epi_growth_mod_treatment))

# stats per species
epi_growth_mod_species = lm(mean_CFU_per_10_leafdiscs ~ species_complex, data = epi_growth_mean_per_rep)

emmeans(epi_growth_mod_species, pairwise ~ species_complex)

hist(resid(epi_growth_mod_species))
```

# Combine epiphytic and virulence data

```{r combine-epi-virulence}
# Combine strain_stats and surv_probs using left_join
epi_virulence_data <- left_join(epi_growth_mean_per_strain, surv_probs, by = "treatment")

# Print the combined data
print(epi_virulence_data)

# Calculate correlation coefficient between survival probability and epiphytic growth ability
correlation <- -1*cor(epi_virulence_data$surv_prob, epi_virulence_data$mean_CFU_per_10_leafdiscs, method = "pearson")
```

# Plot epi virulence plot

```{r epi-virulence-plot}
# Create scatter plot for 72 hours
epi_virulence_plot_strains <- ggplot(epi_virulence_data, aes(x = mean_CFU_per_10_leafdiscs, y = 1-surv_prob)) +
  geom_point(aes(color = species_complex.y)) +
  geom_smooth(method = "lm", se = TRUE, color = "black", fill = "black", alpha = 0.15, linewidth = 0.5) +
  labs(title = "72 Hours",
       x = "Mean CFU per 10 Leaf Discs (Log Transformed)",
       y = "Aphid Mortality at 72 Hours (%)",  
       caption = paste("Pearson's Correlation Coefficient:", round(correlation, 2))) +
  scale_x_continuous(trans = "log10", limits = c(10^2.5, 10^5.5)) +  # Apply log transformation to x-axis
  scale_fill_manual(values = species_colors) +
  scale_color_manual(values = species_colors) +
  theme_minimal()

epi_virulence_plot_strains
```

------------------------------------------------------------------------

# Session Information

```{r session-info}
devtools::session_info()
```
