---
title: "Pseudomonas Epiphytic Growth and Virulence Analysis"
date: "2024 Summer"
output:
  pdf_document:
    toc: true
  html_document:
    css: css/tufte.css
    toc: true
    toc_float: true
  word_document:
    toc: true
    reference_docx: templates/wordstyletemplate.docx
    keep_md: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy.opts = list, tidy = TRUE, fig.align = "center", # Always relative to the document directory
                      fig.path = "../figures/", # Send any figures to this folder
                      dev = "png",  # Export figures as PNG
                      dpi = 300)  # Set the resolution to 300 dots per inch (dpi)
```

# Load Libraries

```{r load-packages}
pacman::p_load(ggplot2, readxl, RColorBrewer, paletteer, viridis, wesanderson, ggbeeswarm, dplyr, tidyverse, devtools, emmeans, cowplot, knitr, survival, here, tibble, survminer, lubridate, formatR, gridExtra, ggsurvfit, gtsummary, tidycmprsk, install = FALSE)

# Load strain colors
# Define your strains
strains <- c(
  "194", "200", "204", "205", "215", 
  "216", "220", "221", "227", "228", 
  "B728a", "Cit7", "Control", "pisi"
)

# Generate colors using the "grDevices::Set 2" palette
set2_colors <- as.character(paletteer_c("grDevices::Dark 3", length(strains)))

# Create a named vector for strain colors, with "Control" as black
strain_colors <- setNames(c(set2_colors), strains)

species_colors <- c(
  "P. fluorescens" = "dodgerblue2",
  "P. syringae" = "springgreen3",
  "P. paralactis" = "sienna4")
```

```{r load-data}
pseud_epi_growth_2024summer_R <- read_excel("data/pseud_epi_growth_2024summer_R.xlsx")

aphid_virulence_data <- read_csv("data/virulence_new_all.csv")
```

# Goals

-   Create Kaplan-Meier curve for Pseud. virulence data
-   Use stats (Wilcox?) to determine statistical significance of each strain
-   Compare virulence data with epiphitic growth ability

# Virulence Analysis

Note: *In order to help me with this analysis, I am using the following sites - [Survival Analysis in R](https://www.emilyzabor.com/tutorials/survival_analysis_in_r_tutorial.html) and [Hazard Ratio: Interpretation & Definition](https://statisticsbyjim.com/probability/hazard-ratio/).*

## Caclulcate survival probabilities for each strain and create dataframe

```{r test-survival}
# Fit the survival model
km_fit <- survfit(Surv(time, censored) ~ treatment, data = aphid_virulence_data)

# Extract survival probabilities at specific time points
time_points <- c(24, 48, 72)
km_summary <- summary(km_fit, times = time_points)

# Initialize empty lists to store the results
times_list <- list()
treatment_list <- list()
surv_prob_list <- list()

# Loop over each treatment group and extract survival probabilities at specified time points
for (i in 1:length(km_fit$strata)) {
  treatment_name <- names(km_fit$strata)[i]
  for (t in time_points) {
    idx <- which(km_summary$time == t & km_summary$strata == treatment_name)
    if (length(idx) > 0) {
      times_list <- c(times_list, t)
      treatment_list <- c(treatment_list, treatment_name)
      surv_prob_list <- c(surv_prob_list, km_summary$surv[idx])
    } else {
      times_list <- c(times_list, t)
      treatment_list <- c(treatment_list, treatment_name)
      surv_prob_list <- c(surv_prob_list, NA)
    }
  }
}

# Create the data frame
surv_probs <- data.frame(
  time = unlist(times_list),
  treatment = unlist(treatment_list),
  surv_prob = unlist(surv_prob_list)
)

# Replace "treatment=" with an empty string
surv_probs$treatment <- gsub("treatment=", "", surv_probs$treatment)

# Print the data frame
print(surv_probs)

# subset data by time
surv_probs_72 <- surv_probs %>% filter(time == 72)

# If you need to save it to a file, you can use the following command
# write.csv(surv_probs, "survival_probabilities.csv", row.names = FALSE)
```

### Post-hoc test for survival

```{r post-hoc}
# Perform pairwise log-rank tests between treatment groups and the control group
pairwise_tests <- pairwise_survdiff(Surv(time, censored) ~ treatment, data = aphid_virulence_data)

# Print the pairwise tests
print(pairwise_tests)
```

### Plot Surival Dataframe at different times

```{r surival-barplot}
surv_plot_72 <- ggplot(data = surv_probs_72, aes(x = treatment, y = (1-surv_prob), color = treatment, fill = treatment)) +
  geom_point() +
  labs(title = "72 hours",
       x = "-80 Strain #",
       y = "Aphid Mortality at 72 Hours (%)") +
  theme_minimal() +
  scale_fill_manual(values = strain_colors) + 
  scale_color_manual(values = strain_colors) +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

surv_plot_72
```

## Make Kaplan-Meier Plot

### Cohort Survival Curve

```{r Kaplan-Meier}
survfit2(Surv(time, censored) ~ treatment, data = aphid_virulence_data) %>% 
  ggsurvfit() +
  labs(
    x = "Hours",
    y = "Overall survival probability"
  ) +
  scale_color_manual(values = strain_colors)
```

# Epiphytic Growth Analysis

## Calculate mean/variance epiphytic growth ability

```{r mean-epi-growth}
# Replace NA with a lower value or remove them for visualization
# Remove rows with NA in CFU_per_10_leafdiscs
epi_growth_clean <- pseud_epi_growth_2024summer_R %>% 
  filter(!is.na(CFU_per_10_leafdiscs))

# Convert CFU_per_10_leafdiscs to numeric, handling scientific notation
epi_growth_clean$CFU_per_10_leafdiscs <- as.numeric(gsub("<", "", epi_growth_clean$CFU_per_10_leafdiscs))

# Subset data to remove unfinished strains and extract the letter part from the 'plant_rep' column
epi_growth_clean <- epi_growth_clean %>%
  filter(!strain %in% c("pisi", "215")) %>%
  mutate(plant_rep_letter = substr(plant_rep, 1, 1))

# Calculate the mean and SD CFU per strain for each plant rep letter and retain the specified columns
epi_growth_mean_per_rep <- epi_growth_clean %>%
  group_by(strain, plant_rep_letter) %>%
  summarise(
    mean_CFU_per_10_leafdiscs = mean(CFU_per_10_leafdiscs, na.rm = TRUE),
    sd_CFU = sd(CFU_per_10_leafdiscs, na.rm = TRUE),
    sample = first(sample),
    person = first(person),
    block = first(block),
    species = first(species),
    plant_rep = first(plant_rep),
    .groups = 'drop'
  )

# Calculate the mean and SD CFU per strain for each plant rep letter and retain the specified columns
epi_growth_mean_per_strain <- epi_growth_clean %>%
  group_by(strain) %>%
  summarise(
    mean_CFU_per_10_leafdiscs = mean(CFU_per_10_leafdiscs, na.rm = TRUE),
    sd_CFU = sd(CFU_per_10_leafdiscs, na.rm = TRUE),
    sample = first(sample),
    person = first(person),
    block = first(block),
    species = first(species),
    plant_rep = first(plant_rep),
    .groups = 'drop'
  )

epi_growth_mean_per_strain
```

## Plot Epiphytic Growth

```{r epi-growth}
epi_growth_strain <- ggplot(data = epi_growth_mean_per_rep, aes(x = strain, y = mean_CFU_per_10_leafdiscs, color = strain, fill = strain)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.3) +
  geom_beeswarm(stroke = 0.5, size = 0.8, alpha = 0.8) +
  labs(title = "CFU per 10 Leaf Discs Across Samples",
       x = "-80 Strain #",
       y = "CFU per 10 Leaf Discs") +
  theme_minimal() +
  scale_fill_manual(values = strain_colors) + 
  scale_color_manual(values = strain_colors) +
  scale_y_log10() +  # Apply log transformation to y-axis
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

epi_growth_strain

epi_growth_species <- ggplot(data = epi_growth_mean_per_rep, aes(x = strain, y = mean_CFU_per_10_leafdiscs, color = species, fill = species)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.3) +
  geom_beeswarm(stroke = 0.5, size = 0.8, alpha = 0.8) +
  labs(title = "CFU per 10 Leaf Discs Across Samples",
       x = "-80 Strain #",
       y = "CFU per 10 Leaf Discs") +
  theme_minimal() +
  scale_fill_manual(values = species_colors) + 
  scale_color_manual(values = species_colors) +
  scale_y_log10() +  # Apply log transformation to y-axis
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "bottom")

epi_growth_species
```

# Combine epiphytic and virulence data

```{r combine-epi-virulence}
# Assuming strain_stats has a column 'strain' and surv_probs has a column 'treatment'
# Rename columns if necessary to match the key for joining
epi_growth_mean_per_strain <- epi_growth_mean_per_strain %>% rename(treatment = strain)

# Combine strain_stats and surv_probs using left_join
epi_virulence_data <- left_join(epi_growth_mean_per_strain, surv_probs, by = "treatment")

# Print the combined data
print(epi_virulence_data)

subset_data_72 <- epi_virulence_data %>% filter(time == 72)

# Calculate correlation coefficient between survival probability and epiphytic growth ability
correlation_72 <- cor(subset_data_72$surv_prob, subset_data_72$mean_CFU_per_10_leafdiscs, method = "pearson")
```

# Plot it for different times

```{r epi-virulence-plot}
# Create scatter plot for 72 hours
epi_virulence_plot_strains <- ggplot(subset_data_72, aes(x = mean_CFU_per_10_leafdiscs, y = (1-surv_prob), color = treatment)) +
  geom_point() +
  geom_smooth(data = subset_data_72, method = "lm", se = FALSE) +
  labs(title = "72 Hours",
       x = "Mean CFU per 10 Leaf Discs (Log Transformed)",
       y = "Aphid Mortality at 72 Hours (%)",  
       caption = paste("Pearson's Correlation Coefficient:", round(correlation_72, 2))) +
  scale_x_continuous(trans = "log10") +  # Apply log transformation to x-axis
  scale_color_manual(values = strain_colors) +
  theme_minimal()

epi_virulence_plot_strains

epi_virulence_plot_species <- ggplot(subset_data_72, aes(x = mean_CFU_per_10_leafdiscs, y = (1-surv_prob), color = species)) +
  geom_point() +
  #geom_smooth(method = "lm", se = FALSE) +
  labs(title = "72 Hours",
       x = "Mean CFU per 10 Leaf Discs (Log Transformed)",
       y = "Aphid Mortality at 72 Hours (%)",  
       caption = paste("Pearson's Correlation Coefficient:", round(correlation_72, 2))) +
  scale_color_manual(values = species_colors) +
  scale_x_continuous(trans = "log10") +  # Apply log transformation to x-axis
    theme_minimal()

epi_virulence_plot_species
```

------------------------------------------------------------------------

# Session Information

```{r session-info}
devtools::session_info()
```
